Morse Theory

### 2020-12-06 paper.tex
Update paper.tex

\begin{equation}
	H(g_1) = \begin{bmatrix} 4 & 0 \\ 0 & 4 \end{bmatrix}
	\imply \det(H(g_1)) = 16
\end{equation}
\begin{equation}
	H(g_2) = \begin{bmatrix} -4 & 0 \\ 0 & -4 \end{bmatrix}
	\imply \det(H(g_2)) =  16
\end{equation}
Both critical points are non-degenerate, therefore $f$ is a Morse function.$\blacksquare$\\
% % end M1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % start V1
\\
\\
\begin{example}
$f(x,y)=e^{xy}+x$ is a Morse function
\end{example}
First of all we have to find all critical points of $f$. A point $P$ is critical, when $\frac{\partial f}{\partial x}=\frac{\partial f}{\partial y}=0$ at $P$. \\
$\frac{\partial f}{\partial x}= ye^{xy}+1=0 \\
\frac{\partial f}{\partial y}=xe^{xy}=0$
After solving the system of equations we get that $x=0$, $y=-1$. Hence, the only critical point of $f$ is $(0,-1)$. Now to prove that the function is a Morse function we have to shoe that the determinant of Hessian matrix at the point $(0,-1)$ is non-zero. \\
Partial second order derivatives are: \\
$\frac{\partial^2 f}{\partial x^2}=y^2e^{xy}$
$\frac{\partial^2 f}{\partial x^2}(0,-1)=1$\\
$\frac{\partial^2 f}{\partial \partial y}=e^{xy}+yxe^{xy}$
$\frac{\partial^2 f}{\partial \partial y}(0,-1)=1$\\
$\frac{\partial^2 f}{\partial y \partial x}=e^{xy}+xye^{xy}$
$\frac{\partial^2 f}{\partial y \partial x}(0,-1)=1$\\
$\frac{\partial^2 f}{\partial y^2}=x^2e^{xy}$
$\frac{\partial^2 f}{\partial y^2}(0,-1)=0$\\
Therefore, \\
$det(Hess(f(0,-1)))=\left| \begin{array}{cc} 1 & 1 \\ 1 & 0 \end{array} \right|=-1$. \\
Since the determinant is non-zero at the only critical point of the function, we can claim that it is a Morse function. \\
\begin{example}
Let's consider a function $g(x,y)=xy$ and prove that it is , indeed, a Morse function.\\

\end{example}
First order partial derivatives: \\
$\frac{\partial g}{\partial x}=y$\\
$\frac{\partial g}{\partial y}=x$\\
Therefore, the only critical point of $g$ is $(0,0)$. Now we have to evaluate the determinant of Hessian matrix at this point.\\
Second order partial derivatives: \\
$\frac{\partial^2 g}{\partial x^2}=0$\\
$\frac{\partial^2 g}{\partial x \partial y}=1$\\
$\frac{\partial^2 g}{\partial y \partial x}=0$\\
$\frac{\partial^2 g}{\partial^2 y}=1$\\
Hence the determinant of Hessian at the critical point is: \\
$det(Hess(f(0,0)))=\left| \begin{array}{cc} 0 & 1 \\ 1 & 0 \end{array} \right|=-1$. \\
We observe that the determinant is nonzero, hence, the only critical point of $g(x,y)=xy$ is non-degenerate, so the function is Morse. \\
\\
\\
\subsection{Examples of non-Morse functions} 
\begin{example}
$f(x,y)=x^3+xY^2-x^2y-y^3$
\end{example}
Let's follow the previous procedure: \\
$\frac{\partial f}{\partial x}=3x^2+y^2-2xy=0$\\
$\frac{\partial f}{\partial y}=2xy-x^2-3y^2=0$\\
After solving the system of equations we got that the only critical point of $f$ is $(0,0)$. Now compute second order partial derivatives:\\
$\frac{\partial^2 f}{\partial x^2}=6x-2y$\\
$\frac{\partial^2 f}{\partial x \partial y}=2y-2x$\\
$\frac{\partial^2 f}{\partial y \partial x}=2y-2x$\\
$\frac{\partial^2 f}{\partial^2 y}=2x-6y$\\
Therefore, \\
$det(Hess(f(0,0)))=\left| \begin{array}{cc} 0 & 0 \\ 0 & 0 \end{array} \right|=0$.\\
Hence, the only critical point of $f$ is degenerate, so $f$ is not a Morse function. \\
$g(x,y)=x^3$\\
$\frac{\partial g}{\partial x}=3x^2$\\
$\frac{\partial g}{\partial y}=0$\\
Therefore, critical points are all points of the form $(0, y_1)$.\\
$\frac{\partial^2 g}{\partial x^2}=6x$
$\frac{\partial^2 g}{\partial x \partial y}=0$\\
$\frac{\partial^2 g}{\partial y \partial x}=0$
$\frac{\partial^2 g}{\partial^2 y}=0$\\
Hence, Hessian will be of the form: \\
$Hess(g(0,y_1))=\left| \begin{array}{cc} 0 & 0 \\ 0 & 0 \end{array} \right|,  \forall y_1$. \\
Therefore, all critical points of $g$ are degenerate points, moreover, values of $g$ at all critical points are equal, hence, $g(x,y)=x^3$ is not a Morse function. 
% % end V1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section{Reeb graphs}
     \subsection{Introduction to Reeb graphs}
     \paragraph{}
      Suppose we have an equivalence relation, $\sim$, defined on a topological space $\X$. Let $\X_\sim$
      be the set of equivalence classes and let $\phi : \X \to \X_\sim$ map each point $x$ to its equivalence class.

>>>>

\begin{equation}
	H(g_1) = \begin{bmatrix} 4 & 0 \\ 0 & 4 \end{bmatrix}
	\imply \det(H(g_1)) = 16
\end{equation}
\begin{equation}
	H(g_2) = \begin{bmatrix} -4 & 0 \\ 0 & -4 \end{bmatrix}
	\imply \det(H(g_2)) =  16
\end{equation}
Both critical points are non-degenerate, therefore $f$ is a Morse function.$\blacksquare$\\
% % end M1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% % start V1
\begin{example}
$f(x,y)=e^{xy}+x$ is a Morse function
\end{example}
First of all we have to find all critical points of $f$. A point $P$ is critical, when at $P$ 
\begin{equation}
\frac{\partial f}{\partial x}=\frac{\partial f}{\partial y}=0\;\;\;
\frac{\partial f}{\partial x}= ye^{xy}+1=0 \;\;\;
\frac{\partial f}{\partial y}=xe^{xy}=0
\end{equation}

After solving the system of equations we get that $x=0$, $y=-1$. Hence, the only critical point of $f$ is $(0,-1)$. Now to prove that the function is a Morse function we have to shoe that the determinant of Hessian matrix at the point $(0,-1)$ is non-zero. \\
Partial second order derivatives are: 
\begin{equation}
 \frac{\partial^2 f}{\partial x^2}=y^2e^{xy}  =>
\frac{\partial^2 f}{\partial x^2}(0,-1)=1    
\end{equation}
\begin{equation}
\frac{\partial^2 f}{\partial \partial y}=e^{xy}+yxe^{xy} =>
\frac{\partial^2 f}{\partial \partial y}(0,-1)=1  
\end{equation}
\begin{equation}
\frac{\partial^2 f}{\partial y \partial x}=e^{xy}+xye^{xy} =>
\frac{\partial^2 f}{\partial y \partial x}(0,-1)=1    
\end{equation}
\begin{equation}
\frac{\partial^2 f}{\partial y^2}=x^2e^{xy}=>
\frac{\partial^2 f}{\partial y^2}(0,-1)=0
\end{equation}

Therefore, 
\begin{equation}
det(Hess(f(0,-1)))=\left| \begin{array}{cc} 1 & 1 \\ 1 & 0 \end{array} \right|=-1
\end{equation}

Since the determinant is non-zero at the only critical point of the function, we can claim that it is a Morse function. \\
\begin{example}
Let's consider a function $g(x,y)=xy$ and prove that it is , indeed, a Morse function.
\end{example}
First order partial derivatives: 
\begin{equation}
\frac{\partial g}{\partial x}=y \;\;
\frac{\partial g}{\partial y}=x 
\end{equation}

Therefore, the only critical point of $g$ is $(0,0)$. Now we have to evaluate the determinant of Hessian matrix at this point.\\
Second order partial derivatives: 
\begin{equation}
\frac{\partial^2 g}{\partial x^2}=0 \;\;
\frac{\partial^2 g}{\partial x \partial y}=1\;\;
\frac{\partial^2 g}{\partial y \partial x}=0\;\;
\frac{\partial^2 g}{\partial^2 y}=1\;\;
\end{equation}
\begin{equation}
\frac{\partial^2 g}{\partial x^2}=0\;\;
\frac{\partial^2 g}{\partial x \partial y}=1\;\;
\frac{\partial^2 g}{\partial y \partial x}=0\;\;
\frac{\partial^2 g}{\partial^2 y}=1
\end{equation}

Hence the determinant of Hessian at the critical point is:
\begin{equation}
det(Hess(f(0,0)))=\left| \begin{array}{cc} 0 & 1 \\ 1 & 0 \end{array} \right|=-1   
\end{equation}

We observe that the determinant is nonzero, hence, the only critical point of $g(x,y)=xy$ is non-degenerate, so the function is Morse. \\
\\
\\
\subsection{Examples of non-Morse functions} 
\begin{example}
$f(x,y)=x^3+xy^2-x^2y-y^3$
\end{example}
Let's follow the previous procedure: 
\begin{equation}
\frac{\partial f}{\partial x}=3x^2+y^2-2xy=0 \;\;
\frac{\partial f}{\partial y}=2xy-x^2-3y^2=0    
\end{equation}

After solving the system of equations we got that the only critical point of $f$ is $(0,0)$. Now compute second order partial derivatives:
\begin{equation}
\frac{\partial^2 g}{\partial x^2}=0 \;\;
\frac{\partial^2 g}{\partial x \partial y}=1 \;\;
\frac{\partial^2 g}{\partial y \partial x}=0 \;\;
\frac{\partial^2 g}{\partial^2 y}=1
\end{equation}

Therefore, 
\begin{equation}
det(Hess(f(0,0)))=\left| \begin{array}{cc} 0 & 0 \\ 0 & 0 \end{array} \right|=0   
\end{equation}

Hence, the only critical point of $f$ is degenerate, so $f$ is not a Morse function. 
\begin{example}
$g(x,y)=x^3$ is not Morse function
\end{example}
\begin{equation}
\frac{\partial g}{\partial x}=3x^2\;\;
\frac{\partial g}{\partial y}=0
\end{equation}

Therefore, critical points are all points of the form $(0, y_1)$.
\begin{equation}
\frac{\partial^2 g}{\partial x^2}=6x \;\;
\frac{\partial^2 g}{\partial x \partial y}=0 \;\;
\frac{\partial^2 g}{\partial y \partial x}=0 \;\;
\frac{\partial^2 g}{\partial^2 y}=0
\end{equation}

Hence, Hessian will be of the form:
\begin{equation}
Hess(g(0,y_1))=\left| \begin{array}{cc} 0 & 0 \\ 0 & 0 \end{array} \right|,  \forall y_1 
\end{equation}

Therefore, all critical points of $g$ are degenerate points, moreover, values of $g$ at all critical points are equal, hence, $g(x,y)=x^3$ is not a Morse function. 
% % end V1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\section{Reeb graphs}
     \subsection{Introduction to Reeb graphs}
     \paragraph{}
      Suppose we have an equivalence relation, $\sim$, defined on a topological space $\X$. Let $\X_\sim$
      be the set of equivalence classes and let $\phi : \X \to \X_\sim$ map each point $x$ to its equivalence class.
